---
title: "Data Collection and Preparation"
output: html_notebook
---

# 1. Install required packages

```{r}
install.packages("tidyverse")
library(tidyverse) # for overall data importing and wrangling
library(lubridate) # for date and time related functions
library(janitor) # for basic cleaning and wrangling functions
library(skimr) # for dataframe summary
```
# 2. Collect data

Upload .csv datasets into the RStudio environment
```{r}
Q22019_Cyclistic <- read.csv("C:/Users/jguic/OneDrive/Desktop/Coding/Google Capstone/Case Study 1 - Cyclistic/Raw Data/Divvy_Trips_2019_Q2/Divvy_Trips_2019_Q2.csv")
Q32019_Cyclistic <- read.csv("C:/Users/jguic/OneDrive/Desktop/Coding/Google Capstone/Case Study 1 - Cyclistic/Raw Data/Divvy_Trips_2019_Q3/Divvy_Trips_2019_Q3.csv")
Q42019_Cyclistic <- read.csv("C:/Users/jguic/OneDrive/Desktop/Coding/Google Capstone/Case Study 1 - Cyclistic/Raw Data/Divvy_Trips_2019_Q4/Divvy_Trips_2019_Q4.csv")
Q12020_Cyclistic <- read.csv("C:/Users/jguic/OneDrive/Desktop/Coding/Google Capstone/Case Study 1 - Cyclistic/Raw Data/Divvy_Trips_2020_Q1/Divvy_Trips_2020_Q1.csv")
```

# 3. Combine data

In order to combine all the dataframes into one, all must share the same columns with the same exact names (although not in the same order)

```{r}
compare_df_cols(Q22019_Cyclistic, Q32019_Cyclistic, Q42019_Cyclistic, Q12020_Cyclistic) # gives an overview of all the columns in the various dataframes
```

At first glance, we're able to see some incongruencies:
  - Q2_2019 and Q3_2019 share column names;
  - Q1_2020 does not have a gender nor a birthday column;
  - Q2_2019, Q3_2019 and Q4_2019 do not have latitude and longitude columns.

## Rename columns

Rename columns to make them consistent with the Q12020 data frame (assuming this will be the chosen table format going forward)
```{r}
head(Q12020_Cyclistic)
```

### Q2_2019

inspect column names
```{r}
colnames(Q22019_Cyclistic)  
```
```{r}
head(Q22019_Cyclistic)
```

```{r}
Q22019_Cyclistic <- Q22019_Cyclistic %>%
  rename(ride_id = "X01...Rental.Details.Rental.ID", 
         rideable_type = "X01...Rental.Details.Bike.ID",
         started_at = "X01...Rental.Details.Local.Start.Time",
         ended_at = "X01...Rental.Details.Local.End.Time", 
         start_station_name = "X03...Rental.Start.Station.Name", 
         start_station_id = "X03...Rental.Start.Station.ID", 
         end_station_name = "X02...Rental.End.Station.Name", 
         end_station_id = "X02...Rental.End.Station.ID",
         member_casual = "User.Type")
```


### Q3_2019

inspect column names
```{r}
colnames(Q32019_Cyclistic)
```

```{r}
Q32019_Cyclistic <- Q32019_Cyclistic %>% 
  rename(ride_id = "trip_id", 
         rideable_type = "bikeid",
         started_at = "start_time",
         ended_at = "end_time", 
         start_station_name = "from_station_name", 
         start_station_id = "from_station_id", 
         end_station_name = "to_station_name", 
         end_station_id = "to_station_id",
         member_casual = "usertype")
```

### Q4_2019

inspect column names
```{r}
colnames(Q42019_Cyclistic)
```

```{r}
Q42019_Cyclistic <- Q42019_Cyclistic %>% 
  rename(ride_id = "trip_id", 
         rideable_type = "bikeid",
         started_at = "start_time",
         ended_at = "end_time", 
         start_station_name = "from_station_name", 
         start_station_id = "from_station_id", 
         end_station_name = "to_station_name", 
         end_station_id = "to_station_id",
         member_casual = "usertype")
```

## Change data types

As noticed previously, all the ride_id and ride_type columns must be formatted to the character data type in order to be appended
```{r}
compare_df_cols(Q22019_Cyclistic, Q32019_Cyclistic, Q42019_Cyclistic, Q12020_Cyclistic)
```

```{r}
Q22019_Cyclistic$rideable_type <- as.character(Q22019_Cyclistic$rideable_type)
Q32019_Cyclistic$rideable_type <- as.character(Q32019_Cyclistic$rideable_type)
Q42019_Cyclistic$rideable_type <- as.character(Q42019_Cyclistic$rideable_type)

Q22019_Cyclistic$ride_id <- as.character(Q22019_Cyclistic$ride_id)
Q32019_Cyclistic$ride_id <- as.character(Q32019_Cyclistic$ride_id)
Q42019_Cyclistic$ride_id <- as.character(Q42019_Cyclistic$ride_id)
```

## Append dataframes

```{r}
Y_Cyclistic <- bind_rows(Q22019_Cyclistic, Q32019_Cyclistic, Q42019_Cyclistic, Q12020_Cyclistic)
```

## Delete unnecessary columns

```{r}
Y_Cyclistic <- Y_Cyclistic %>% 

  select(-c(start_lat,
            start_lng,
            end_lat,
            end_lng,
            birthyear,
            gender,
            X01...Rental.Details.Duration.In.Seconds.Uncapped,
            X05...Member.Details.Member.Birthday.Year,
            tripduration,
            Member.Gender
            )
         )
```

# 4. Final overview
```{r}

skim_without_charts(Y_Cyclistic)
```

