---
title: "Data Cleaning and Manipulation"
output: html_notebook
---



# 1. Data Inspection and Planning

Let's inspect the data
```{r}
summary(Y_Cyclistic)
```

```{r}
head(Y_Cyclistic)
```

The main problems that need to be fixed are:

- remove unwanted characters and duplicate entries
- fix inconsistent data
- include new calculated time columns
- remove maintenance trips, zero and negative time trips

# 2. Data Cleaning

## 2.1 Remove unwanted characters and duplicate entries

The clean_names function will remove any unwanted characters and consolidate data to facilitate aggregation
```{r}
clean_names(Y_Cyclistic)
```

Let's look for duplicate entries

```{r}
Y_Cyclistic %>% 
  distinct() %>% 
  nrow()
```
Since the number of distinct values is equal to the total number of observations in the data frame, we conclude that all rows are unique

## 2.2 Creating column consistency

Knowing that the member_casual column has different names for members ("member" and "Subscriber") and casual riders ("Customer" and "casual"), we need to consolidate them from 4 to 2 labels
```{r}
table(Y_Cyclistic$member_casual)
```

Replace the Column values using the recode() function
```{r}
Y_Cyclistic <- Y_Cyclistic %>%
  mutate(member_casual = recode(member_casual,
                                "Subscriber" = "member",
                                "Customer" = "casual"
                                )
         )
```

Confirm that all values have been corrected
```{r}
table(Y_Cyclistic$member_casual)
```

## 2.3 Create new columns

We can create day, weekday, month and Year columns to aggregate data in different levels of granularity and thus facilitating the discovery of new trends

```{r}

Y_Cyclistic$date <- as.Date(Y_Cyclistic$started_at) # we can directly create a new date column
Y_Cyclistic$month <-format(as.Date(Y_Cyclistic$date), "%m")
Y_Cyclistic$day <-format(as.Date(Y_Cyclistic$date), "%d")
Y_Cyclistic$year <-format(as.Date(Y_Cyclistic$date), "%Y")
Y_Cyclistic$weekday <-format(as.Date(Y_Cyclistic$date), "%A")

skim_without_charts(Y_Cyclistic)
```

We can also go further with out possibilities and create start_hour and ride_length column
```{r}
Y_Cyclistic$start_hour <- hour(Y_Cyclistic$started_at)

Y_Cyclistic$ride_length <- difftime(strptime(Y_Cyclistic$ended_at, "%Y-%m-%d %H:%M:%S"), #ended_at needs to be first converted to time data type
                                    strptime(Y_Cyclistic$started_at, "%Y-%m-%d %H:%M:%S") #ended_at needs to be first converted to time data type
                                    )
```

Since the ride_length column is formatted to "x seconds", we want to give it a cleaner look and format it to "%h:%M:$s"
```{r}
#Load and use the hms package
library(hms)

Y_Cyclistic$ride_length <- as_hms(Y_Cyclistic$ride_length)
```

## 2.4 Remove maintenance and negative time trips

The dataframe includes a few entries when bikes were taken out of docks and checked for quality by Cyclistic or ride_length was negative
```{r}
Y_Cyclistic %>% 
  filter(ended_at == started_at)
```

It is noticible that the start_station_name values appear to all be HQ QR. We're able to confirm it
```{r}
Y_Cyclistic %>%
  filter(ended_at == started_at) %>% 
  count(start_station_name) #groups by start_station_name and counts the number of observations 
```

Now let's cross the information and see if all HQ QR trips have zero ride_length
```{r}
Y_Cyclistic %>%
  filter(start_station_name == "HQ QR") %>% 
  count(ride_length) 
```
As some HQ QR trips have positive ride lengths, we're only able to use the ended_at == started_at filter condition

```{r}
Y_Cyclistic %>% 
  filter(ride_length < 0)
```

So we now have to remove observations where:
  - "HQ QR" start_station_names has they are maintenance trips
  - ride_length is equal or less than zero
  
Since data is being removed, the best practice is to create and work with a new dataframe
```{r}
Y_Cyclistic_V2 <- Y_Cyclistic[!(Y_Cyclistic$start_station_name == "HQ QR" | Y_Cyclistic$ride_length <= 0),]

nrow(Y_Cyclistic) - nrow(Y_Cyclistic_V2)
```
As we can see, 3780 observations have been deleted
